---
title: "Module 2: Visualisation"
author: "Stefano Cacciatore"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
  toc: true
  toc_depth: 2
  theme: united
  toc_float:
      collapsed: true
---

# Visualisation
Data visualisation helps to tell stories about the data in a form that is appealing and easy to understand and draw conclusions. It is said that "a picture is worth a thousand words"

R is equipped with powerful predefined visualisation features() and more add-on features used to create complex but aesthetically pleasing graphs  

**Learning Objectives:**

1. Understand and Apply Scatter Plots

2. Analyze and Interpret Box Plots

3. Create and Customize Heatmaps

4. Conduct and Visualize Survival Analysis

#### Tidyverse packages

Tidyverse is a collection of packages designed for data analysis and includes:

- `dplyr` for data frames manipulations

- `purrr` for functional programming

- `ggplot2` for data visualizations

- `readr` for reading various types of data files

installing the tidyverse package and loading the library

```
#install.packages("tidyverse")
library(tidyverse)
```


```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
```
## 1. Scatter Plots

Scatter plots are useful for visualizing the relationship between two continuous variables.


```{r scatter-plot}
library(ggplot2)
library(dplyr)

head(starwars)
#summary(starwars)

# cleaning the dataset to remove NA values for height & mass 
starwars_QC <- starwars %>%
  filter(!is.na(height)&!is.na(mass))

# filtering the dataset to remove outliers
starwars_QC1 <- starwars_QC %>%
  filter(height < 300 & mass < 500)

# Create scatter plot
ggplot(starwars_QC1, aes(x = height, y = mass)) +
  geom_point() +
  labs(title = "Relationship between Mass and Height", x = "Height(cm)", y = "Mass(kg)") +
  theme_minimal()
```



#### Correlation coefficient.

The correlation between 2 variables is found with the `cor()` function.

Suppose we want to compute the correlation between height and mass in the `starwars_QC1` dataset:

```{r scatter-plot-correlation}
# Calculate correlation
correlation <- cor(starwars_QC1$height, starwars_QC1$mass, use = "complete.obs")
print(paste("Correlation coefficient between height and mass:", round(correlation, 2)))
```

The `Pearson` correlation is computed by default with the cor() function. If you want to compute the `Spearman` or `Kendall's tau-b` correlation, add the argument `method = "spearman"`


```{r linear_regression}
# Fit linear model
model <- lm(mass ~ height, data = starwars_QC1)

# Print model summary
summary(model)
```
Plotting the regression line for the relationship

``` {r regression_plot}
ggplot(starwars_QC1, aes(x = height, y = mass)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(title = "Relationship between Height and Mass",
       x = "Height (cm)",
       y = "Mass (kg)") +
  theme_minimal()
```


```{r regresion_diagnostics}
# Regression diagnostics
par(mfrow = c(2, 2))  # Set up plotting area for 4 plots
plot(model)
```

## 2. Box plots
A boxplot or a box-and-whisker plot is a standard method used to display the distribution of a dataset based on its five-number summary of data points; `minimum`, `first quartile[Q1]`, `median`, `third quartile[Q3]`, and `maximum`.

A box plot can tell you that your data is symmetrical, has outliers or skewed
``` {r iris_dataset}
data("iris")
head(iris)
```
Create a box plot of Sepal Length by Species using the iris dataset
``` {r simple_boxplot}
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot() +
  labs(title = "Box Plot of Sepal Length by Species",
       x = "Species",
       y = "Sepal Length (cm)") +
  theme_minimal()
```
We can compare the species list
```{r ggboxplot}
#install.packages("ggpubr")
library(ggpubr)
library(ggplot2)

# Define the colors for the boxplot
cols <- c("red", "blue", "green")


# Prepare the dataset (using Sepal.Length as the continuous variable)
df <- data.frame(Sepal.Length = iris$Sepal.Length, Species = iris$Species)

# Define comparisons for pairwise Wilcoxon test
my_comparisons <- list(c("setosa", "versicolor"), c("setosa", "virginica"), c("versicolor", "virginica"))

# Create the boxplot
ggboxplot(df, x = "Species", y = "Sepal.Length", width = 0.8, palette = cols,
          fill = "Species", add = "jitter", ylim = c(4, 9.5),  # Increase ylim to provide space for labels
          add.params = list(size = 0.5, jitter = 0.2, fill = "Species"),
          shape = 21) +
  labs(y= "Sepal Length", x = "Species", title = "Box plot sepal length by species") +
  stat_compare_means(method = "kruskal.test") +  # Overall comparison using Kruskal-Wallis test
  stat_compare_means(comparisons = my_comparisons, 
                     method = "wilcox.test", 
                     label = "p.signif", 
                     label.y = c(8.5, 9.2, 8.7)) +  # Manually adjust the position of significance labels
  theme(legend.position = "none", plot.margin = unit(c(2, 1, 1, 1), "cm"))
```

## 3. Heatmaps

Heat maps or heatmaps allow us to simultaneously visualize clusters of samples and features. 

There are multiple R packages and functions for drawing interactive and static heatmaps, including:

- `heatmap()` [R base function, stats package]: Draws a simple heatmap

- `heatmap.2()` [gplots R package]: Draws an enhanced heatmap compared to the R base function.

- `pheatmap()` [pheatmap R package]: Draws pretty heatmaps and provides more control to change the appearance of heatmaps.

- `d3heatmap()` [d3heatmap R package]: Draws an interactive/clickable heatmap

- `Heatmap()` [ComplexHeatmap R/Bioconductor package]: Draws, annotates and arranges complex heatmaps (very useful for genomic data analysis)

We shall use the `mtcars` dataset. 
```{r heatmap}
#standardising the data to make variables comparable
df <- scale(mtcars)
# Default plot
heatmap(df, scale = "none")
```
We can customise by using the `RColorBrewer` color palette to change the appearance.
The arguments `RowSideColors` and `ColSideColors` are used to annotate rows and columns respectively. 
```{r heatmap_custom}
library("RColorBrewer")
col <- colorRampPalette(brewer.pal(10, "RdYlBu"))(256)
heatmap(df, scale = "none", col =  col, 
        RowSideColors = rep(c("blue", "pink"), each = 16),
        ColSideColors = c(rep("purple", 5), rep("orange", 6)))
```



## 4. Survival plots

Survival analysis involves a set of statistical approaches used to investigate the time it takes for an event of interest to occur. 

In cancer studies, survival analysis is used for patients survival time analyses with typical research questions like;

- What is the impact of certain clinical characteristics on patient’s survival?

- What is the probability that an individual survives after 3 years?

- Are there differences in survival between groups of patients?

```{r survival, warning=FALSE}
#install.packages('survminer')
BiocManager::install("RTCGA.clinical") # data for examples

library(survminer)
library(RTCGA.clinical)
survivalTCGA(BRCA.clinical, OV.clinical,
             extract.cols = "admin.disease_code") -> BRCAOV.survInfo

library(survival)
fit <- survfit(Surv(times, patient.vital_status) ~ admin.disease_code,
               data = BRCAOV.survInfo)

# Visualize with survminer
ggsurvplot(fit, data = BRCAOV.survInfo, risk.table = TRUE)
```
The simple plot above shows an estimate of survival probability depending on days from cancer diagnostics grouped by cancer types.
Additionally an informative risk set table which shows the number of patients that were under observation in the specific period of time.

One can create a powerful informative survival plot with such specification of the parameters
```{r survival_plot_customised}
ggsurvplot(
  fit,                     # survfit object with calculated statistics.
  data = BRCAOV.survInfo,  # data used to fit survival curves. 
  risk.table = TRUE,       # show risk table.
  pval = TRUE,             # show p-value of log-rank test.
  conf.int = TRUE,         # show confidence intervals for point estimates of survival curves.
  xlim = c(0,2000),        # present narrower X axis, but not affect survival estimates.
  break.time.by = 500,     # break X axis in time intervals by 500.
  ggtheme = theme_minimal(), # customize plot and risk table with a theme.
  risk.table.y.text.col = T, # colour risk table text annotations.
  risk.table.y.text = FALSE # show bars instead of names in text annotations in legend of risk table
)
```
