---
title: "Data Analysis in R: Module 1"
author: "Stefano Cacciatore"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
  toc: true
  toc_depth: 2
  theme: united
  toc_float:
      collapsed: true
---

# Basics in R

##### Learning Objectives:

-   Familiarize with the RStudio interface.
-   Learn basic R Operations.
-   How to understand and create functions.
-   Dealing with packages.
-   Importing external data.
-   Basic data manipulation.

#### 

> ***Open RStudio and take a few minutes to explore each pane and its functionality.***

### 1. Familiarize with the RStudio Interface

In this section, we'll get to know the RStudio interface and understand its different components:

-   **Layout**: The RStudio interface consists of several panes that help you manage your workflow efficiently.

-   **Panes**: These include the Source Pane, Console, Environment/History Pane, and the Files/Plots/Packages/Help/Viewer Pane.

-   **Console**: The Console is where you can directly execute R commands.

-   **Script Editor**: The Script Editor is used for writing and editing scripts, which can be executed in parts or as a whole.

-   **Environment**: The Environment Pane shows all the objects (data, functions, etc.) that you've created during your session.

### 2. Basic R Operations

*Let's start by performing some basic operations in R:*

**Arithmetic:**

-   addition, subtraction, multiplication, division etc.

```{r}
# Basic Arithmetic
sum <- 5 + 3 
difference <- 5 - 3
product <- 5 * 3
quotient <- 5 / 3
exponent <- 5 ^ 3
modulus <- 5 %% 3 # Remainder from division
int_division <-5 %/% 3 # Integer Division

# Display results
sum
difference
product
quotient
exponent
modulus
int_division
```

**Variable Assignment:**

-   You can assign values to variables for reuse.

```{r}
# Variable Assignment
x <- 10 
y <- 20
z <- x + y # x & y are now saved in your environment for reuse
z
```

**Data Types:** - Include Numeric, Character, Boolean, Integer, and Double types of data.

```{r}
n <- numeric(6)
typeof(n) # a function used to return the type of the data stored
```

```{r}
c <- "c"
typeof(c)
```

```{r}
# Integer values are whole numbers that can be positive or negative
i <- integer(5)
typeof(i)
```

```{r}
# Boolean Values consists of data hat has one of two values (e.g., 1 or 0 / True or False)
t <- TRUE # can store as letter "T"
f <- FALSE # can store as letter "F"
typeof(t)
typeof(f) # logical = boolean 
```

**Data Formats:**

-   vectors,

-   lists,

-   data frames

```{r}

# Vectors are arrays of data elements each of the SAME TYPE.
vec <- c(1, 2, 3, 4, 5)

# Lists contain multiple items that can be of different data types like numbers characters and also could contain stored vectors or data frames.
lst <- list(name = "John", age = 25)

# Data Frames are a tabular (2-dimensional) data structure that can store values of any data type.
df <- data.frame(Name = c("Alice", "Bob"), Age = c(25, 30))

vec
lst
df
```

[***Exercise 1***: Try creating your own variables and performing different operations. Experiment with different data types.]{style="color:red;"}

[***Exercise 2***: Use the function `str(x)` to check the structure of each data type.]{style="color:red;"}

<details>

<summary>Answer</summary>

```{r, echo=T}
str(vec)
str(lst)
str(df)
```

### 3. Functions in R

Understanding and creating functions is fundamental in R:

-   Functions are reusable blocks of code that perform a specific task.

Creating Simple Functions: Here's how you can create a simple function in R.

```{r}
# Creating a Function
add_numbers <- function(a, b) {  
  return(a + b)  # takes two arguments (a and b) and returns their sum.
}

# Using the Function: Here we are calling the function 'add_numbers' with 10 and 15 as inputs.
add_numbers(10, 15)
```

[***Exercise 3***: Create a function that takes a number as input and returns the square of that number.]{style="color:red;"}

<details>

<summary>Answer</summary>

```{r, echo=TRUE}
sq <- function(n1,n2) {
  first_step <- n1 + n2
  return(first_step^2) # also can use the function sqrt(first_step)
}

sq(2,3)
```

### 4. Installing and loading Packages:

-   R packages are collections of functions and datasets developed by the R community:

-   There are pre-loaded packages in Rstudio that can be used and called without installation (e.g., dplyr)

-   CRAN: CRAN (Comprehensive R Archive Network) is the main repository for R packages.

Installing and Loading Packages: To use additional functions, you might need to install and load packages.

```{r}

# Installing a Package (Uncomment the line below if the package has been loaded previously)
#     install.packages("ggplot2")

# Loading the package "ggplot" a data visualtion package we will use later in the course:
library(ggplot2) # no quotations needed when loading a package from your library
ggplot

```

[***Exercise 4***: Install and load the dplyr package. Use it to manipulate a data frame of your choice.]{style="color:red;"}

***Hint*** **use the code `browseVignettes(package = "dplyr")` to see how you could use this package.**

<details>

<summary>Answer</summary>

```{r, echo=TRUE}
#     install.packages("dplyr")
#     ?dplyr
#     browseVignettes(package = "dplyr") --> see what you can do with dplyr
library(dplyr)

# Manipulating data:

data <- data.frame(
  Name = c("Alice", "Bob", "Charlie", "David", "Eva"),  # Example data frame
  Age = c(23, 35, 45, 29, 31),
  Score = c(89, 72, 95, 88, 76)
) 

# 1. Select specific columns (Name and Score)
selected_data <- data %>%
  select(Name, Score)

# 2. Filter rows where Age is greater than 30
filtered_data <- data %>%
  filter(Age > 30)

# 3. Arrange the data by Score in descending order
arranged_data <- data %>%
  arrange(desc(Score))

```

```{r, echo=FALSE}
# View the results
print("Selected Data:")
print(selected_data)

print("Filtered Data (Age > 30):")
print(filtered_data)

print("Arranged Data by Score (Descending):")
print(arranged_data)
```

### 5. Importing External Data:

-   Data analysis often involves importing data from external files:

-   R can read data from various formats like CSV, Excel, etc.

-   It is important to pay attention to the extension the file uses (e.g., csv = comman seperated values)

```{r}

# Reading CSV Files:
# df_csv <- read.csv("path/to/your/file.csv")

# Reading Excel Files (requires the readxl package)
#   install.packages("readxl)
#   library(readxl)
# df_excel <- read_excel("path/to/your/file.xlsx")

```

***TIP:*** If you look at the top left of Rstudio you can import the dataset manually using the "Import Dataset" option.

[***Exercise 5***: Try reading a sample CSV file into R. Practice using different data formats if available.]{style="color:red;"}

<details>

<summary>Answer</summary>

```{r, echo=TRUE}
# df_csv <- read.csv("path/to/sample/file/sample.csv", sep = ",")
# df_table <- read.table("path/to/sample/file/sample.tsv, sep = "\t"")
# df_table <- read.excel("path/to/sample/file/sample.xlsx)

# If file is a text (.txt) file, remember to check what the extension is before reading it in. 
```

### 6. Indexing:

-   I have stored a data frame (with example data not shown) named "df"

```{r, echo=FALSE}
# Can store columns of different data types

city <- c("Cairo", "Kinshasa", "Lagos", "Luanda", "Dar es Salaam", "Khartoum",
          "Johannesburg", "Abidjan", "Addis Ababa", "Nairobi", "Yaoundé", 
          "Casablanca", "Antananarivo", "Kampala", "Kumasi" , "Dakar", 
          "Ouagadougou", "Lusaka", "Algiers", "Bamako", "Brazzaville",
          "Mogadishu", "Tunis", "Conakry", "Lomé", "Matola", "Monrovia",
          "Harare", "N'Djamena", "Nouakchott", "Niamey", "Freetown", 
          "Lilongwe", "Kigali", "Abomey-Calavi", "Tripoli", "Bujumbura",
          "Asmara", "Bangui", "Libreville")

abb <- c("CA", "KI", "LA", "LU", "DS", "KH", "JO", "AB", "AD", "NA", "YA",
         "CB", "AN", "KA", "KU", "DA", "OU", "LS", "AL", "BA", "BR", "MO",
         "TU", "CO", "LO", "MA", "MN", "HA", "ND", "NO", "NI", "FR", "LI", 
         "KG", "AC", "TR", "BU", "AS", "BG", "LB")

region <- c("North", "Central", "Central", "South", "South", "North",
            "South", "Central", "Central", "Central", "Central",
            "North", "South", "Central", "Central", "Central",
            "Central", "South", "North", "North", "South",
            "Central", "North", "Central", "Central", "South",
            "Central", "South", "North", "North", "North",
            "Central", "South", "Central", "Central", "North",
            "South", "Central", "Central", "Central")

region <- factor(region, levels = c("North", "Central" , "South"))

pop <- c(22183200, 16315534, 15387639, 8952496, 7404689, 
         6160327, 6065354, 5515790, 5227794, 5118844, 
         4336670, 3840396, 3699900, 3651919, 3630326, 
         3326001, 3055788, 3041789, 2853959, 2816943, 
         2552813, 2497463, 2435961, 2048525, 1925517, 
         1796872, 1622582, 1557740, 1532588, 1431539, 
         1383909, 1272145, 1222325, 1208296, 1188736, 
         1175830, 1139265, 1034872, 933176, 856854)

total <- c(125700, 80500, 66900, 51700, 36400, 45700, 31000, 
           37600, 41300, 28600, 24600, 25000, 9300, 11800,
           14200, 23200, 21900, 32100, 29300, 9700, 5300,
           6500, 13500, 20700, 35100, 11600, 3600, 11100,
           9700, 2100, 12000, 9300, 6300, 2200, 8400, 
           6700, 2700, 3200,1200, 700)

```

```{r}
df <- data.frame(city, abb, region, pop, total)
head(df) # shows the first 'n' rows
```

-   Indexing is the process of selecting elements using their indices (i.e., positions in the data format)

*Indexing ROWS and COLUMNS by POSITION:*

```{r}
# Select 1st Row:
first_row <- df[1, ]  # Selects the entire first row
first_row

# Select 2nd column:
second_column <- df[, 2]  # Selects the entire second column ('abb')
second_column

# Select the element in the 3rd row and 4th column:
specific_element <- df[3, 4]  # Selects the population ('pop') of the third city ('Lagos')
specific_element
```

*Indexing Using COLUMN Names:*

```{r}
# Select the 'city' column:
city_column <- df$city  # Selects the 'city' column
city_column

# Select the first row using column names:
first_row_city_pop <- df[1, c("city", "pop")]  # Selects the 'city' and 'pop' columns for the first row
first_row_city_pop

# Logical Indexing:
large_cities <- df[df$pop > 10000000, ]  # Returns all rows where 'pop' is greater than 10 million
large_cities
```

*Indexing with `dplyr`:*

```{r}
# Using dplyr you can perform similar operations with clearer syntax:
library(dplyr)

# Select specific columns
selected_df <- df %>%
  select(city, pop)  # Selects 'city' and 'pop' columns

selected_df

# Filter rows based on a condition
filtered_df <- df %>%
  filter(pop > 10000000)  # Filters for cities with population greater than 10 million

filtered_df
```

[***Exercise 6***: Given the following dataframe]{style="color:red;"}

```{r}
df <- data.frame(
  city = c("Cairo", "Kinshasa", "Lagos", "Luanda", "Dar es Salaam"),
  abb = c("CA", "KI", "LA", "LU", "DS"),
  region = factor(c("North", "Central", "Central", "South", "South"), 
                  levels = c("North", "Central", "South")),
  pop = c(22183200, 16315534, 15387639, 8952496, 7404689),
  total = c(125700, 80500, 66900, 51700, 36400)
)

head(df)
```

[a. Select the Population of the First City (Cairo):]{style="color:red;"}

[b. Select the Abbreviation for the Third City]{style="color:red;"}

[c. Select the Entire Row for the Fourth City]{style="color:red;"}

[d. Select the 'city' and 'pop' Columns for the Second and Fifth Cities]{style="color:red;"}

[e. Filter for Cities in the 'South' Region]{style="color:red;"}

[f. Using dplyr, Select the city and total Columns]{style="color:red;"}

<details>

<summary>Answer</summary>

```{r, echo=T}
# a. Population of the First City
df[1, "pop"]

# b. Abbreviation for the Third City
df[3, "abb"]

# c. Entire Row for the Fourth City
df[4, ]

# d. 'city' and 'pop' Columns for the Second and Fifth Cities
df[c(2, 5), c("city", "pop")]

# e. Filter for Cities in the 'South' Region
df[df$region == "South", ]

# f. Using dplyr, Select the 'city' and 'total' Columns
library(dplyr)
df %>%
  select(city, total)

```

### 7. Data Manipulation Basics

Basic data manipulation is key to preparing data for analysis.

***Overview of Data Classes: Learn about data frames, matrices, and lists.***

-   Data Frames: Rectangular tables with rows and columns, where columns can be of different types.

-   Matrices: Rectangular tables with rows and columns, where all elements must be of the same type.

-   Lists: Collections of elements that can be of different types, including other lists.

```{r}
# Creating a Data Frame
df <- data.frame(
  city = c("Cairo", "Kinshasa", "Lagos", "Luanda", "Dar es Salaam"),
  population = c(22183200, 16315534, 15387639, 8952496, 7404689),
  region = c("North", "Central", "Central", "South", "South")
)
head(df)

# Creating a Matrix
matrix_example <- matrix(1:9, nrow = 3, byrow = TRUE)
head(matrix_example)

# Creating a List
list_example <- list(
  numbers = 1:5,
  text = c("A", "B", "C"),
  data_frame = df
)
print(list_example)
```

-   Subsetting Data: Extract subsets of data based on conditions.

```{r}
# Extract cities with population greater than 10 million
large_cities <- df[df$population > 10000000, ]
print(large_cities)

# Extract cities in the 'South' region
south_cities <- df[df$region == "South", ]
print(south_cities)

# Select Specific Columns
city_population <- df[, c("city", "population")]
print(city_population)
```

-   Basic Transformations: Perform simple data transformations such as adding new columns.

```{r}
# Add a New Column: City Area (dummy data)
df$area <- c(606, 851, 1171, 300, 400)  # Example areas in square kilometers
print(df)

# Modify Existing Column: Increase Population by 10%
df$population <- df$population * 1.10
print(df)

# Using dplyr for Data Manipulation:

# Add a New Column: City Area
df <- df %>%
  mutate(area = c(606, 851, 1171, 300, 400))

# Modify Existing Column: Increase Population by 10%
df <- df %>%
  mutate(population = population * 1.10)

# Filter: Cities in the 'South' Region
south_cities_dplyr <- df %>%
  filter(region == "South")

# Select Specific Columns
city_population_dplyr <- df %>%
  select(city, population)
```

```{r}
print(df)
print(south_cities_dplyr)
print(city_population_dplyr)
```

<span style="color:red;">***Exercise 7***:<span>

<span style="color:red;">a. Data Classes: Extract the population of the 3rd city from the df data frame.<span>

<span style="color:red;">b. Subsetting Data -\> Extract the elements from matrix_example that are greater than 5.<span>

<span style="color:red;">c. Basic Transformations: Add a new column area to df with values 500, 600, 700, 800, 900.<span>

<details>

<summary>Answer</summary>

```{r, echo=TRUE}
# 7a:
df[3, "population"]
```

```{r, echo=TRUE}
# 7b:
matrix_example[matrix_example > 5]
```

```{r, echo=TRUE}
# 7c:
df$area <- c(500, 600, 700, 800, 900)
head(df)
```
