---
title: "Module 3: Univariate Analysis"
author: "Stefano Cacciatore"
output: 
  html_document:
  toc: true
  toc_depth: 2
  theme: united
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---

## Univariate Analysis

*What is univariate analysis ?*

The idea of univariate analysis is to first understand the variables individually. It is typically the first step in understanding a data set. A variable in UA is a condition or subset that your data falls into. You can think of it as a “category" such as "age", "weight" or "length". However, UA does not look at \> than 1 variable at a time (this would be a bivariate analysis)

#### Learning Objectives:

-   Summarising Data

-   Frequency Tables

-   Univariate Hypothesis Testing

-   Visualising Univariate Data

- Correlation

- Simple Regression analysis


```{r, message=FALSE, warning=FALSE}
# Installation of packages (usually needed)
# install.packages("ggplot2")
# install.packages("dplyr")
# install.packages("ggpubr")
# install.packages("corrplot")

# Loading of packages
library(ggplot2)
library(dplyr)
library(ggpubr)
library(corrplot)
library(stats)
```

### 1. Summarising Data

```{r}
data("iris")
# We need to create an array of our variable for Sepal.Length:
x <- iris$Sepal.Length
```

Looking at the **CENTRAL TENDENCY** of the data:

- `mean(x)`

- `median(x)`

- `mode(x)`


Looking at the **DISPERSION** of the data:

- `min(x)`

- `max(x)`

- `range(x)` - Range of the data

- `IQR(x)`- Inter-quantile range

- `var(x)` - Variance

- `sd(x)` - Standard Deviation


TIP: you can use the function `summary` for a quick look at your data

```{r}
summary(iris)
```

#### Frequency Tables:

-   The frequency of an observation tells you the number of times the observation occurs in the data.

-   A frequency table is a collection of these observations and their frequencies.

-   A frequency table can be shown either graphically (bar chart/histogram) or as a frequency distribution table.

-   These tables can show qualitative (categorical) or quantitative (numeric) variables.

```{r}
# Frequency table for the categorical variable
freq_table <- table(iris$Species)
freq_table
```

**Note:** the frequency table is CASE-SENSITIVE so the frequencies of the variables corresponds to how many times that specific number of string appears.

### 2. Correlation:

Correlation analysis is used to investigate the association between two or more variables.

#### Step 1: Choose a Correlation Method

***Pearson Correlation*** measures the linear relationship between two continuous variables. It assumes both variables follow a normal distribution.

***Spearman and Kendall Correlation*** are non-parametric and measure the strength and direction of the association between two ranked variables.

#### Step 2: Calculate Correlation

i. Pearson Correlation

```{r}
# Pearson correlation test
pearson_res <- cor.test(mtcars$mpg, mtcars$wt, method = "pearson")
pearson_res
```


ii. Spearman and Kendall Correlation

```{r, warning=FALSE}
# Spearman correlation test
spearman_res <- cor.test(mtcars$mpg, mtcars$wt, method = "spearman")
spearman_res

# Kendall correlation test
kendall_res <- cor.test(mtcars$mpg, mtcars$wt, method = "kendall")
kendall_res
```

#### Step 3: Interpretation

Correlation Coefficient:

- `-1`: Strong negative correlation (as one variable increases, the other decreases).

- `0`: No correlation.

- `1`: Strong positive correlation (both variables increase together).

P-Value:

- `p-value < 0.05` indicates a statistically significant correlation.

[***Exercise:***]{style="color:red;"}

Perform a correlation analysis using the `mpg` and `qsec` variables from the `mtcars` to investigate the extent of correlation between the two variables. Provide an interpretation of the correlation coefficient and its p-value. 

<details>

<summary>*Example interpretation*:</summary>

- The Pearson correlation coefficient is -0.8677, which points to a strong negative linear relationship between the variables. 

- The p-value is significantly low (p < 0.001), indicating that the correlation is statistically significant. 

- The 95% confidence interval suggests that the true correlation lies between -0.9338 and -0.7441.

### 3. Simple Linear Regression:

```{r}
x <- mtcars$mpg
y <- mtcars$wt

model = lm(y ~ x)
summary(model)
```

```{r}
par(mfrow = c(2, 2))
plot(model)
```

#### *Checking Assumptions*:

##### **Assumption 1: Linearity** --> Check if the relationship between variables is linear.

Plot of `x` vs `y`: This scatter plot displays the relationship between the predictor `x` and the response `y`.

Abline (Regression Line): The `abline(model)` adds the fitted regression line to the plot.

```{r}
plot(x, y)
abline(model)
```

*What to Look For*:

**Linear Relationship**: The data points should roughly form a straight line if the linearity assumption is satisfied. The fitted regression line should capture the trend of the data points well.

**Non-Linearity**: If the data points show a clear curvature or systematic pattern not captured by the straight line, this suggests that the linearity assumption is violated. In such cases, consider polynomial regression or other non-linear models.


##### **Assumption 2: Homoscedasticity** --> Ensure that the residuals are evenly distributed.

*Plot of Residuals*: This plot shows the residuals from the model. Residuals are the differences between the observed values and the values predicted by the model.

```{r}
plot(model$residuals)
```

*What to Look For*:

**Even Spread**: Ideally, the residuals should be randomly scattered around zero and should not display any clear pattern. This indicates homoscedasticity (constant variance of residuals).

**Patterns**: If you observe a pattern, such as a funnel shape (residuals increasing or decreasing as `x` increases), it suggests heteroscedasticity (non-constant variance). In such cases, consider transforming the dependent variable or using robust regression techniques.

##### **Assumption 3: Normality of Residuals** --> Use Q-Q plots to check the normality of residuals.

Q-Q Plot: The Q-Q plot (quantile-quantile plot) compares the quantiles of the residuals with the quantiles of a normal distribution.

```{r}
qqnorm(model$residuals)
qqline(model$residuals)
```

*What to Look For*:

**Straight Line**: If the residuals are normally distributed, the points should closely follow the straight line (`qqline`). This suggests that the normality assumption is reasonable.

**Deviations**: Significant deviations from the line indicate that the residuals are not normally distributed. This could mean the presence of outliers or skewness in the residuals. If the normality assumption is violated, consider transforming the response variable or using non-parametric methods.

##### ***Summary of Interpretation***:

- **Linearity**: The plot of x vs. y with the regression line should show a clear linear relationship.

- **Homoscedasticity**: The plot of residuals should display no obvious patterns or systematic structures.

- **Normality of Residuals**: The Q-Q plot should show residuals following the diagonal line if they are normally distributed.

These plots help you validate the assumptions underlying your regression model, ensuring that your results are reliable and interpretable.

[***Simple Linear Regression Exercise:***]{style="color:red;"}

You were asked to analyze the following dataset, mtcars, where mpg (miles per gallon) is used as the predictor variable and wt (weight) as the response variable. You have fitted a linear regression model and checked the assumptions.

Now, perform simple linear regression on two variables of your choosing from the mtcars data set and answer the following questions:

1. Describe the relationship between mpg and wt. Does the plot suggest a linear relationship?

2. Describe the spread of the residuals. Is there any noticeable pattern that might suggest a violation of the homoscedasticity assumption?

3. Assess whether the residuals appear to follow a normal distribution based on the Q-Q plot. Are there any significant deviations from the diagonal line?


