---
title: "Single-cell pipeline"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---



## Single-cell data


For this tutorial, we will be analyzing the a dataset of Peripheral Blood Mononuclear Cells (PBMC) freely available from 10X Genomics. There are 2,700 single cells that were sequenced on the Illumina NextSeq 500. The raw data can be found [here](https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz).


We start by reading in the data. The Read10X() function reads in the output of the cellranger pipeline from 10X, returning a unique molecular identified (UMI) count matrix. The values in this matrix represent the number of molecules for each feature (i.e. gene; row) that are detected in each cell (column). Note that more recent versions of cellranger now also output using the h5 file format, which can be read in using the Read10X_h5() function in Seurat.

We next use the count matrix to create a Seurat object. The object serves as a container that contains both data (like the count matrix) and analysis (like PCA, or clustering results) for a single-cell dataset. For more information, check out our [Seurat object interaction vignette], or our GitHub Wiki. For example, in Seurat v5, the count matrix is stored in pbmc[["RNA"]]$counts.

```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}
library(dplyr)
library(Seurat)
library(patchwork)
library(KODAMA)
library(KODAMAextra)


# Load the PBMC dataset
pbmc.data <- Read10X(data.dir = "../singlecell_tutorial/filtered_gene_bc_matrices/hg19/")
# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc
```

QC and selecting cells for further analysis
Seurat allows you to easily explore QC metrics and filter cells based on any user-defined criteria. A few QC metrics commonly used by the community include

The number of unique genes detected in each cell.
Low-quality cells or empty droplets will often have very few genes
Cell doublets or multiplets may exhibit an aberrantly high gene count
Similarly, the total number of molecules detected within a cell (correlates strongly with unique genes)
The percentage of reads that map to the mitochondrial genome
Low-quality / dying cells often exhibit extensive mitochondrial contamination
We calculate mitochondrial QC metrics with the PercentageFeatureSet() function, which calculates the percentage of counts originating from a set of features
We use the set of all genes starting with MT- as a set of mitochondrial genes
# The [[ operator can add columns to object metadata. This is a great place to stash QC stats

```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

```

#Where are QC metrics stored in Seurat?
The number of unique genes and total molecules are automatically calculated during CreateSeuratObject()
You can find them stored in the object meta data
```
# Show QC metrics for the first 5 cells
head(pbmc@meta.data, 5)
```


In the example below, we first visualize key quality control (QC) metrics and then apply thresholds to remove low-quality or outlier cells.

We retain only cells that meet the following criteria:

Have between 200 and 2,500 detected features (genes)

Have more than 500 total RNA counts

Contain less than 5% mitochondrial transcripts, which helps exclude stressed or dying cells

```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}
# Visualize QC metrics as a violin plot
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```


```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}
pbmc <- subset(pbmc, subset = nCount_RNA>500 & nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
```


#Normalizing the data
After removing unwanted cells from the dataset, the next step is to normalize the data. By default, we employ a global-scaling normalization method “LogNormalize” that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result. In Seurat v5, Normalized values are stored in pbmc[["RNA"]]$data.

```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}
pbmc <- NormalizeData(pbmc)
```
While this method of normalization is standard and widely used in scRNA-seq analysis, global-scaling relies on an assumption that each cell originally contains the same number of RNA molecules. We and others have developed alternative workflows for the single cell preprocessing that do not make these assumptions. For users who are interested, please check out our SCTransform() normalization workflow. The method is described in ourpaper, with a separate vignette using Seurat here. The use of SCTransform replaces the need to run NormalizeData, FindVariableFeatures, or ScaleData (described below.)



Identification of highly variable features (feature selection)
We next calculate a subset of features that exhibit high cell-to-cell variation in the dataset (i.e, they are highly expressed in some cells, and lowly expressed in others). We and others have found that focusing on these genes in downstream analysis helps to highlight biological signal in single-cell datasets.

Our procedure in Seurat is described in detail here, and improves on previous versions by directly modeling the mean-variance relationship inherent in single-cell data, and is implemented in the FindVariableFeatures() function. By default, we return 2,000 features per dataset. These will be used in downstream analysis, like PCA.

```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}

pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

```

Scaling the data
Next, we apply a linear transformation (‘scaling’) that is a standard pre-processing step prior to dimensional reduction techniques like PCA. The ScaleData() function:

Shifts the expression of each gene, so that the mean expression across cells is 0
Scales the expression of each gene, so that the variance across cells is 1
This step gives equal weight in downstream analyses, so that highly-expressed genes do not dominate
The results of this are stored in pbmc[["RNA"]]$scale.data
By default, only variable features are scaled.
You can specify the features argument to scale additional features

```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)
```

#Perform linear dimensional reduction
Next we perform PCA on the scaled data. By default, only the previously determined variable features are used as input, but can be defined using features argument if you wish to choose a different subset (if you do want to use a custom subset of features, make sure you pass these to ScaleData first).

For the first principal components, Seurat outputs a list of genes with the most positive and negative loadings, representing modules of genes that exhibit either correlation (or anti-correlation) across single-cells in the dataset.
```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}

pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
DimPlot(pbmc, reduction = "pca")
```
#Clustering on PCA and UMAP visualization
Build a KNN graph in PCA space and perform graph-based clustering; the resolution parameter controls granularity (higher → more clusters). Then compute a UMAP using the selected PCs to obtain a 2D visualization that preserves local neighborhoods—use this plot to review cluster separations and potential mixing.

```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}
pbmc <- FindNeighbors(pbmc, dims=1:2)
pbmc <- FindClusters(pbmc, resolution = 0.3)
pbmc <- RunUMAP(pbmc, reduction = "pca", dims = 1:20)
DimPlot(pbmc, reduction = "umap")
```

#KODAMA analysis


```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}
pbmc=RunKODAMAmatrix(pbmc,reduction="pca",dims=20,n.cores=4)
pbmc=RunKODAMAvisualization(pbmc)
```

#Clustering on KODAMA

```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}

pbmc <- FindNeighbors(pbmc, dims=1:2, reduction="KODAMA")
pbmc <- FindClusters(pbmc, resolution = 0.05)
DimPlot(pbmc, reduction = "KODAMA")
```
#Parameters adjustment

```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}

config=umap.defaults
config$n_neighbors=50
pbmc=RunKODAMAvisualization(pbmc,config=config)

pbmc <- FindNeighbors(pbmc, dims=1:2, reduction="KODAMA")
pbmc <- FindClusters(pbmc, resolution = 0.05)

DimPlot(pbmc, reduction = "KODAMA")
```
#Find markers
Identifies differentially expressed genes between two groups of cells using a Wilcoxon Rank Sum test (default); will use a fast implementation by Presto if installed
```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}

#remotes::install_github("immunogenomics/presto")
markers <- FindAllMarkers(pbmc)
head(markers)
```








```{r,fig.width=10,fig.height=8,warning=FALSE,message=FALSE}
markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
DoHeatmap(pbmc, features = top10$gene) + NoLegend()
```

```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}
markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 5) %>%
    ungroup() -> top5

DotPlot(
  pbmc,
  features = unique(top5$gene),
  group.by = "seurat_clusters",
  assay = "RNA"
) + RotatedAxis()
```


```{r,fig.width=10,fig.height=4,warning=FALSE,message=FALSE}

VlnPlot(pbmc, features = c("MS4A1", "CD79A"))
```



```{r,fig.width=10,fig.height=8,warning=FALSE,message=FALSE}

FeaturePlot(pbmc, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A", "FCGR3A", "LYZ", "PPBP", "CD8A"),reduction = "KODAMA")

temp=as.vector(pbmc$RNA_snn_res.0.05)
temp[temp=="0"]="T cells memory"
temp[temp=="1"]="Naive T cells"
temp[temp=="2"]="Monocytes"
temp[temp=="3"]="B cells"
temp[temp=="4"]="CD8 T effector"
temp[temp=="5"]="NK cells"
temp[temp=="6"]="CD16+ monocytes"
temp[temp=="7"]="CD8 T effector-memory"
pbmc$cell_type=as.factor(temp)  
Idents(pbmc) <- pbmc$cell_type

DimPlot(pbmc, reduction = "KODAMA")
```


Naive CD4+ T: IL7R, CCR7	
CD14+ Mono: CD14, LYZ	
Memory CD4+: IL7R, S100A4	
B cell: MS4A1	
CD8+ T cell: CD8A	
FCGR3A+ Monocyte: FCGR3A, MS4A7	
NK cell: GNLY, NKG7	
DC: FCER1A, CST3	
Platelet: PPBP	

```{r,fig.width=10,fig.height=8,warning=FALSE,message=FALSE}

# Install (only once)
if (!requireNamespace("remotes")) install.packages("remotes")
remotes::install_github("saezlab/liana")

library(liana)
library(magrittr)

liana_res <- liana_wrap(pbmc)


liana_agg <- liana_res %>%
  liana_aggregate()

# Quick check
head(liana_agg)
# columns include: source, target, ligand.complex, receptor.complex, aggregate_rank, etc.

# 2) Dotplot for cluster 0 -> clusters 1 and 2
liana_agg %>%
  liana_dotplot(
    source_groups = c("T cells memory"),         # sender
    target_groups = c("Naive T cells", "Monocytes","B cells"),    # receivers
    ntop = 20
  ) + RotatedAxis()
```
